---
title: "Validating traverse against the historical workflow"
description: "Quantitative equivalence of passage calculations"
format:
  gfm:
    toc: true
    toc-depth: 2
    code-fold: show
    code-summary: "Show code"
    preview-mode: raw
editor: source
execute:
  warning: false
  message: false
---

`traverse` was extracted from a working landscape-connectivity workflow. This
note validates the documented historical configuration against that original
implementation before the package is generalized to broader applications.

## Why validate equivalence?

This is software/model equivalence validation, not a claim that the historical
workflow was scientifically optimal. The comparison holds the landscape,
conductance values, terminal coordinates, pairings, transition settings,
passage settings, and target mask constant. The historical implementation is
run directly with `raster`, `sp`, and `gdistance`; the new implementation keeps
those classes behind the `traverse` API.

```{r}
#| label: setup
#| include: false
pkgload::load_all(".", quiet = TRUE)
library(terra)
library(ggplot2)
library(patchwork)
library(traverse)

validation_env <- new.env(parent = globalenv())
options(traverse.skip_historical_validation = TRUE)
sys.source(
  file.path("scripts", "validate-historical-nws.R"),
  envir = validation_env
)
options(traverse.skip_historical_validation = FALSE)
validation <- validation_env$run_historical_validation(
  root = normalizePath(".", mustWork = TRUE),
  n_nodes = 2L,
  write_figures = FALSE
)
template <- validation$template
raw_surface <- validation$raw_surface
domain <- validation$domain
surface <- validation$surface
sources <- validation$sources
targets <- validation$targets
pairs <- validation$pairs
```

## Shared landscape and parameters

The repaired repository assets provide a controlled proxy for the historical
application. Finite cells in `template.tif` define the target study area, and
the surface has no missing values in those finite target cells. Missing cells
outside the target are assigned the same low conductance in both workflows.

```{r}
#| label: settings-table
knitr::kable(
  validation$settings_table,
  caption = "Inputs and settings held constant between implementations."
)
```

The target raster is extended by a 150 map-unit computational buffer. The
surface is transformed as `x / 100`, given a finite minimum conductance of
`0.01`, and assigned `0.01` in the non-target computational area. Two source
and two target terminals create four source-target passage pairs. The
validation function accepts a larger `n_nodes` value when a full 3 x 3 run is
practical on the host system.

## Shared terminals

The terminal coordinates are generated once with `traverse_nodes()` and passed
unchanged to both implementations. This is important: independently
regenerating "equivalent" terminals could introduce a geometry difference that
would be mistaken for a model difference.

```{r}
#| label: terminals-figure
#| fig-cap: "The same source and target coordinates are supplied to the historical and traverse implementations. The red rectangle is the target extent; the grey rectangle is the computational extent."
validation$plots$terminals
```

## Historical implementation

The reference path converts the prepared conductance surface to a legacy
`RasterLayer`, creates a 16-neighbour transition layer, applies
`geoCorrection(type = "c", multpl = FALSE, scl = TRUE)`, calculates passage for
each of the four pairs with `theta = 1e-100`, averages those pairwise rasters,
and masks the aggregate back to the target.

```{r}
#| label: historical-code
#| eval: false
legacy_raster <- raster::raster(surface$raster)
raster::crs(legacy_raster) <- terra::crs(surface$raster, proj = TRUE)
legacy_transition <- gdistance::transition(
  legacy_raster, transitionFunction = mean, directions = 16
)
legacy_transition <- gdistance::geoCorrection(
  legacy_transition, type = "c", multpl = FALSE, scl = TRUE
)

source_sp <- sp::SpatialPoints(
  terra::crds(sources),
  proj4string = sp::CRS(terra::crs(sources, proj = TRUE))
)
target_sp <- sp::SpatialPoints(
  terra::crds(targets),
  proj4string = sp::CRS(terra::crs(targets, proj = TRUE))
)
legacy_pairwise <- lapply(seq_len(nrow(pairs)), function(i) {
  gdistance::passage(
    legacy_transition,
    origin = source_sp[pairs$source_index[i], , drop = FALSE],
    goal = target_sp[pairs$target_index[i], , drop = FALSE],
    theta = 1e-100
  )
})
legacy_mean <- raster::calc(
  raster::stack(legacy_pairwise),
  fun = function(x) if (all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE)
)
legacy_target <- raster::mask(
  raster::crop(legacy_mean, raster::raster(domain$target_template)),
  raster::raster(domain$target_mask)
)
```

## `traverse` implementation

The equivalent `traverse` call receives the same prepared surface, domain,
terminals, pair table, and model settings. The package converts to legacy
backend classes only inside its transition and passage adapters.

```{r}
#| label: traverse-code
#| eval: false
traverse_result <- traverse(
  surface,
  domain,
  sources = sources,
  targets = targets,
  pairs = pairs,
  directions = 16,
  theta = 1e-100,
  correction = "c",
  scale_correction = TRUE,
  aggregation = "mean"
)
```

The development validation script runs both code paths and fails if the core
acceptance criteria are not met. It also records elapsed time for each path;
runtime is informational here and is not used to claim an optimization.

## Spatial comparison

The historical and `traverse` maps use identical scale limits. The difference
map is signed (`traverse - historical`) and uses a restrained diverging scale
centered at zero.

```{r}
#| label: passage-comparison
#| fig-cap: "Historical and traverse target-domain passage surfaces shown on identical colour scales."
validation$plots$passage
```

```{r}
#| label: difference-map
#| fig-cap: "Signed cell-by-cell difference between the traverse and historical target-domain passage surfaces."
validation$plots$difference
```

## Numerical equivalence

Raster geometry includes dimensions, resolution, extent, CRS, and the finite
target mask. Values are compared only where both outputs are finite, with
relative metrics excluding reference values effectively equal to zero.

```{r}
#| label: raster-summary
knitr::kable(
  validation$summary_table,
  digits = 8,
  caption = "Target-domain raster summaries."
)
```

```{r}
#| label: equivalence-metrics
metric_names <- c(
  "pearson", "spearman", "rmse", "mae",
  "maximum_absolute_difference", "mean_absolute_difference",
  "normalized_rmse", "relative_rmse", "relative_maximum_difference",
  "percent_cells_within_tolerance"
)
metric_table <- validation$metrics_table[
  match(metric_names, validation$metrics_table$Metric), , drop = FALSE
]
knitr::kable(
  metric_table,
  digits = 12,
  caption = "Cell-by-cell equivalence metrics."
)
```

```{r}
#| label: geometry-summary
geometry_table <- data.frame(
  Check = c("Geometry identical", "Finite target mask identical", "Acceptance criteria"),
  Result = c(
    validation$comparison$geometry_identical,
    validation$comparison$finite_mask_identical,
    validation$comparison$acceptance
  )
)
knitr::kable(geometry_table, caption = "Structural equivalence checks.")
```

## Cell-by-cell comparison

The scatter plot provides a direct check of the one-to-one relationship between
the two output rasters. The dashed line is the identity line.

```{r}
#| label: scatter
#| fig-cap: "Historical passage values versus traverse passage values for every finite target cell."
validation$plots$scatter
```

## Interpretation

```{r}
#| label: interpretation
interpretation <- if (isTRUE(validation$comparison$acceptance)) {
  paste(
    "For this documented historical configuration, traverse reproduces the",
    "reference gdistance passage workflow within the reported numerical",
    "tolerance. Geometry and finite target masks are identical, and the",
    "cell-by-cell residuals are within tolerance. This conclusion is limited",
    "to the tested conductance preparation, terminal geometry, pair weighting,",
    "and gdistance settings."
  )
} else {
  paste(
    "The configured equivalence criteria were not met. Inspect the numerical",
    "metrics and signed difference map before changing package behavior."
  )
}
interpretation
```

The historical NWS application motivates this validation, but no NWS-specific
transformation is embedded in the package API. For the documented historical
configuration, the package architecture has now been checked against the
legacy implementation while retaining explicit control over conductance,
terminal placement, transition settings, geographic correction, and pair
aggregation.
